Developing effective theories for the square lattice Hubbard model using density matrix downfolding

### 2019-06-14 draft.tex
finished title page

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{subfigure}
\usepackage[margin=1.0in]{geometry}
\usepackage{float}
\usepackage{ulem}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{wrapfig}
\renewcommand{\baselinestretch}{1.5}

\begin{document}
\section{Introduction}
A core problem of condensed matter physics is understanding how to systematically develop low-energy effective models for strongly correlated electron systems. In the forefront of modern research are materials like high temperature superconductors and heavy fermion systems for which simple effective theories have not yet been found. For example, the cuprate family of high-Tc superconductors have an intricate phase diagram in the doping-temperature space  with a robust anti-ferromagnetic (AFM) charge-transfer insulator state near zero doping, a prominent superconducting dome in the 0.1 - 0.2 hole doped region at low temperatures with psuedo-gap and non-Fermi liquid metallic phases at higher temperatures and Fermi liquid like behavior in the overdoped area, of which only the AFM insulating and Fermi liquid phases have well understood mechanism and models. The iron-pnictides also have superconducting domes, non-Fermi liquid like metallic phases, a spin-density-wave antiferommagnetic phase, all under various dopings and temperatures. Heavy-fermion materials have a purported quantum critical point between an AFM and Fermi-liquid phase which houses a superconducting dome, leading to another phase diagram for which building model Hamiltonians is very challenging.

@@

When downfolding $\hat{H}_\text{ab}$ we will use FN-DMC to access states in $\mathcal{LE}$ and accurately calculate their \textit{ab-initio} energies and RDM elements.
Diffusion Monte Carlo (DMC) is a quantum Monte Carlo method which projects out the ground state of a Hamiltonian given some initial trial wave function.
Consider a trial wave function $|\Psi_T\rangle$ and the Hamiltonian $\hat{H}_\text{ab}$ with ground state $|\Phi_0\rangle$. We apply the projector $e^{-\tau \hat{H}_\text{ab}}$ as $\tau \rightarrow \infty$ to $|\Psi_T \rangle$
\begin{equation}
\lim_{\tau \rightarrow \infty} e^{-\tau \hat{H}_\text{ab}} |\Psi_T\rangle 
\equiv \lim_{\tau \rightarrow \infty} |\Psi_\text{DMC}(\tau)\rangle \propto \langle \Phi_0|\Psi_T\rangle |\Phi_0\rangle,
\end{equation}
projecting out the ground state as long as the trial wave function we choose is not orthogonal to the ground state. 
The stochastic implementation involves moving samples from the trial function $\Psi_T(R)$ using the Green function $G(R, R^\prime, \tau) = \langle R | e^{-\tau(\hat{H}_\text{ab} - E_T)} | R^\prime \rangle$. Since $H_\text{ab} = T + V$, kinetic and potential energy terms, the Green function is approximated by a Trotter expansion $G(R, R^\prime, \tau) = \langle R | e^{-\tau(\hat{H} - E_T)} | R^\prime \rangle \sim \Big[e^{-d\tau(V(R) + V(R^\prime))/2} \langle R| e^{-d\tau(\hat{T} - E_T)}|R^\prime \rangle + O(d\tau^2) \Big]^N $ where $d\tau = \tau/N$.
This expansion can be interpreted as an interative procedure where samples are moved N times with a small timestep $d\tau$ until convergence.
The constant $E_T$ is a trial energy used to control the normalization of $\Psi_\text{DMC}(\tau, R)$ and is updated at each move.
This implementation suffers from a fermion sign problem which is dealt with through a fixed-node approximation where the nodal surface of $\Psi_\text{DMC}(\tau, R)$ is forced to match that of the initial trial wave function for all $\tau$.
This approximation makes FN-DMC variational, and will only return the exact ground state of $\hat{H}_\text{ab}$ if the nodal surfaces of $|\Psi_T\rangle$ and $|\Phi_0\rangle$ are identical.
We make use of this variational principle to access low-energy states within $\mathcal{H}$ by choosing trial wave functions with nodes that differ from the nodes of $|\Phi_0 \rangle$.
To calculate the expectation value of an operator $\hat{A}$ in FN-DMC we will use a mixed estimator $\langle \Psi_{DMC}(\tau) |\hat{A} | \Psi_T \rangle/\langle \Psi_\text{DMC}(\tau) | \Psi_T \rangle$ which allows us to use the importance sampled distribution $f = \Psi_\text{DMC}(\tau)\Psi_T$ for our Monte Carlo estimates.
Details for calculating reduced density matrix elements in FN-DMC can be found in Wagner.

>>>>

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{subfigure}
\usepackage[margin=1.0in]{geometry}
\usepackage{float}
\usepackage{ulem}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{wrapfig}
\renewcommand{\baselinestretch}{1.5}
\author{Shivesh Pathak, Lucas Wagner}
\title{Developing effective theories for the square lattice Hubbard model using density matrix downfolding}

\begin{document}
\maketitle
\begin{abstract}
A core problem of condensed matter physics is understanding how to systematically develop low-energy effective models for strongly correlated electron systems. 
We believe that a density matrix downfolding (DMD) method can be used to systematically build effective theories for such systems beginning with a high energy theory like the \textit{ab-initio} Hamiltonian H$_\text{ab}$.
We present as a preliminary result an effective theory developed using DMD which can accurately reproduce the spectra and properties of the lowest lying excitations of the CuO molecule.
As a step towards bulk strongly correlated materials, we propose developing effective theories using DMD for the simplest toy system which exhibits strongly correlated behavior: the 1-band square lattice Hubbard model for various couplings U, temperatures T and dopings $\delta$. 
\end{abstract}
\pagebreak

\section{Introduction}
A core problem of condensed matter physics is understanding how to systematically develop low-energy effective models for strongly correlated electron systems. 
In the forefront of modern research are materials like high temperature superconductors and heavy fermion systems for which simple effective theories have not yet been found. 
For example, the cuprate family of high-Tc superconductors have an intricate phase diagram in the doping-temperature space  with a robust anti-ferromagnetic (AFM) charge-transfer insulator state near zero doping, a prominent superconducting dome in the 0.1 - 0.2 hole doped region at low temperatures with psuedo-gap and non-Fermi liquid metallic phases at higher temperatures and Fermi liquid like behavior in the overdoped area, of which only the AFM insulating and Fermi liquid phases have well understood mechanism and models. 
The iron-pnictides also have superconducting domes, non-Fermi liquid like metallic phases, a spin-density-wave antiferommagnetic phase, all under various dopings and temperatures. 
Heavy-fermion materials have a purported quantum critical point between an AFM and Fermi-liquid phase which houses a superconducting dome, leading to another phase diagram for which building model Hamiltonians is very challenging.

@@

When downfolding $\hat{H}_\text{ab}$ we will use fixed-node diffusion Monte Carlo (FN-DMC) to access states in $\mathcal{LE}$ and accurately calculate their \textit{ab-initio} energies and RDM elements.
Diffusion Monte Carlo (DMC) is a quantum Monte Carlo method which projects out the ground state of a Hamiltonian given some initial trial wave function.
Consider a trial wave function $|\Psi_T\rangle$ and the Hamiltonian $\hat{H}_\text{ab}$ with ground state $|\Phi_0\rangle$. We apply the projector $e^{-\tau \hat{H}_\text{ab}}$ as $\tau \rightarrow \infty$ to $|\Psi_T \rangle$
\begin{equation}
\lim_{\tau \rightarrow \infty} e^{-\tau \hat{H}_\text{ab}} |\Psi_T\rangle 
\equiv \lim_{\tau \rightarrow \infty} |\Psi_\text{DMC}(\tau)\rangle \propto \langle \Phi_0|\Psi_T\rangle |\Phi_0\rangle,
\end{equation}
projecting out the ground state as long as the trial wave function we choose is not orthogonal to the ground state. 
The stochastic implementation involves moving samples from the trial function $\Psi_T(R)$ using the Green function $G(R, R^\prime, \tau) = \langle R | e^{-\tau(\hat{H}_\text{ab} - E_T)} | R^\prime \rangle$. Since $H_\text{ab} = T + V$, kinetic and potential energy terms, the Green function is approximated by a Trotter expansion $G(R, R^\prime, \tau) = \langle R | e^{-\tau(\hat{H} - E_T)} | R^\prime \rangle \sim \Big[e^{-d\tau(V(R) + V(R^\prime))/2} \langle R| e^{-d\tau(\hat{T} - E_T)}|R^\prime \rangle + O(d\tau^2) \Big]^N $ where $d\tau = \tau/N$.
This expansion can be interpreted as an interative procedure where samples are moved N times with a small timestep $d\tau$ until convergence.
The constant $E_T$ is a trial energy used to control the normalization of $\Psi_\text{DMC}(\tau, R)$ and is updated at each move.
This implementation suffers from a fermion sign problem which is dealt with through a fixed-node approximation where the nodal surface of $\Psi_\text{DMC}(\tau, R)$ is forced to match that of the initial trial wave function for all $\tau$.
This approximation makes FN-DMC variational, and will only return the exact ground state of $\hat{H}_\text{ab}$ if the nodal surfaces of $|\Psi_T\rangle$ and $|\Phi_0\rangle$ are identical.
We make use of this variational principle to access low-energy states within $\mathcal{H}$ by choosing trial wave functions with nodes that differ from the nodes of $|\Phi_0 \rangle$.
To calculate the expectation value of an operator $\hat{A}$ in FN-DMC we will use a mixed estimator $\langle \Psi_{DMC}(\tau) |\hat{A} | \Psi_T \rangle/\langle \Psi_\text{DMC}(\tau) | \Psi_T \rangle$ which allows us to use the importance sampled distribution $f = \Psi_\text{DMC}(\tau)\Psi_T$ for our Monte Carlo estimates.
Details for calculating reduced density matrix elements in FN-DMC can be found in Wagner.
